%\subsection{Equivalence} % (fold)
\label{sec:the_proof}

  Let $P\subset \R^d$ be a set of $n$ points.
  Pick any \emph{source} point $s\in P$.
  Order the points of $P$ as $p_1,\ldots ,p_n$ so that
  \[
    \dist_2(s,p_1) \le \cdots \le \dist_2(s, p_n).
  \]
  This will imply that $p_1 = s$.
  It will suffice to show that for all $p_i\in P$, we have $\dist_2(s,p_i) = \dist_N(s,p_i)$.
  There are three main steps:
  \begin{enumerate}
    \item We first show that when $P$ is a subset of the vertices of an axis-aligned box, $\dist = \dist_N$.  In this case, shortest paths for $\dist$ are single edges and shortest paths for $\dist_N$ are straight lines.
    \item We then show how to lift the points from $\R^d$ to $\R^n$ by a Lipschitz map $m$ that places all the points on the vertices of a box and preserves $\dist_2(s,p)$ for all $p\in P$.
    \item Finally, we show how the Lipschitz extension of $m$ is also Lipschitz as a function between nearest neighbor metrics.  We combine these pieces to show that $\dist \le \dist_N$.  As $\dist \ge \dist_N$ (Lemma~\ref{lem:dist_N_le_dist}), this will conclude the proof that $\dist = \dist_N$.
  \end{enumerate}
The key to the second step, to be elaborated in Section~\ref{sec:lifting},
is that if you take points on a line and raise the pairwise distances to
the $1/2$ power, you get points on a box. This is a special case of the
general theory on screw functions developed by Von Neumann and Schoenberg,
which asserts a far more general criterion on when functions applied to
pairwise distances between points on a line can be embedded into Euclidean
space~\cite{VonNeumann41}.
  \input{screw/boxes-soda}
  \input{screw/lifting-soda}
  \input{screw/lip_extension-soda}
\subsection{From Finite Sets to Finite Collections of Compact Path-Connected Bodies}
\label{sec:bodies}
All of our proof steps hold for finite collections of compact,
path-connected bodies in arbitrarily large dimension. Our Lipschitz map $m$ can
still be extended to a Lipschitz map $f$ in this setting, largely due to the
generality of the Kirszbraun theorem. In this case, the pre-image of the
contractive map is the set of all points belonging to some body. Meanwhile, the image is a finite set of points, the corners of a multi-dimensional box.
Thus our construction of $m$ contracts each convex
body into a single point, and the image of our
compact bodies under $f$ is still a finite point set on the corners of a
box. Therefore, the remainder of our theorem proof goes through unchanged.

This result is rather remarkable: path-connected compact sets in high
dimensional space can have extremely convoluted geometry, and the Voronoi diagrams
on these collections (on which the nearest neighbor metric depends) can be
massively complex.  The key is that our Lipschitz map is robust enough to
handle objects of considerable geometric complexity.
  
  
% section the_proof (end)
